---
title: "MATH 4323 UFO Sightings Project" 
output: html_notebook
---

Contents: 
Data loading: Entire team
Cleaning-null values: 
Cleaning-redundant features, outliers: 
Cleaning-converting categorical/ordinal values to numeric: 
PCA: 
Correlation plots, box plots: 
Distribution plots, word cloud: 
K-means clustering: 
Hierarchical clustering: 

Load the data into a dataframe, replace all blanks and NA with NaN, strip whitespace
```{r}
library(tidyverse)
data <- read.csv("scrubbed.csv", na.strings=c(""," ","NA"), strip.white=TRUE)

```

View the data, 80332 obs. of  11 variables
```{r}
summary(data)
str(data)
levels(data$state)
levels(data$country)
```

View obserations by country
```{r}
summary(data$country)
```


Drop duration in hours and minutes since we already have duration in seconds and date posted
```{r}
data = subset(data, select = -c(city, comments, duration..hours.min., date.posted))
names(data)
```
Check number of NA and clean
```{r}
# is.na(data)
sum(is.na(data))
data<-na.omit(data)
sum(is.na(data))
```

Convert datetime
```{r}
#splitting datetime
library(lubridate)
library(dplyr)
#make sure datetime is formatted correctly to be split
#make separate columns for year, month, day, and time in hours
data <- mutate(data, datetime = mdy_hm(datetime), year_sighted = year(datetime), month_sighted = month(datetime), day_sighted = day(datetime), hour_sighted = hour(datetime))
data
```

Boxplots for observations by time of day, month, year
```{r}
#time of day
boxplot(data$hour_sighted, horizontal=TRUE, main="Time of day UFOs are seen")
#time of month
boxplot(data$day_sighted, horizontal=TRUE, main="Time of month (by day) UFOs are seen")
#time of year
boxplot(data$month_sighted, horizontal=TRUE, main="Time of year (by month) UFOs are seen")
```

```{r}
#time of day
d <- density(data$hour_sighted)
plot(d, type="n", main="UFO reports by time of day (hrs)")
polygon(d, col="green", border="gray")
#time of month
d <- density(data$day_sighted)
plot(d, type="n", main="UFO reports by day of the month")
polygon(d, col="green", border="gray")
#time of year
d <- density(data$month_sighted)
plot(d, type="n", main="UFO reports by month")
polygon(d, col="green", border="gray")
```

Scale the data and convert it to a numeric matrix, 66516, 14 columns
```{r}
numeric_matrix <-scale(data.matrix(data))
# numeric_matrix
summary(numeric_matrix)
str(numeric_matrix)
```

Correlation plot
```{r}
library(corrplot)
corrplot(cor(numeric_matrix), method="color", addCoef.col = "black",  tl.col="black", tl.srt=45)
```

PCA 
```{r}
library(factoextra)

res.pca <-princomp(numeric_matrix, cor = FALSE, scores = TRUE)
summary(res.pca)
```
```{r}
fviz_eig(res.pca)

```


Hierarchical clustering with complete, average, single linkage
```{r}
library(dendextend)
df_sample<-sample_n(data, 100)
sample <-scale(data.matrix(df_sample))

```

Hierarchical clustering with complete linkage
```{r}
fviz_nbclust(sample, FUN = hcut, method = "silhouette", nstart=20)
```


```{r}
complete <- hclust(dist(sample))
plot(complete)
rect.hclust(complete , k = 2, border = 2:8)

```


```{r}
single <- hclust(dist(sample), method = "single")
plot(single)
rect.hclust(single , k = 2, border = 2:8)
```


```{r}
average <- hclust(dist(sample), method = "average")
plot(average)
rect.hclust(average , k = 2, border = 2:8)
```


k-means clustering
```{r}
# determine optimal number of clusters
fviz_nbclust(sample, kmeans, method = "silhouette")
# ok that said k=1 so im ignoring it
```


```{r}
# k-means cluster
km.out <- eclust(kclust_data, FUNcluster="kmeans", k=3, nstart=20)

```

```{r}

plot(kclust_data, col=km.out$cluster, main="K-Means Clustering Results with K=3", pch=20, cex=2)
```

